\vspace{1.0cm} \Large \bf Supplemental material of ``Health and disease imprinted in the time variability of the human microbiome''

### 2016-07-26 mSystems/mSys_MAIN.tex
release version 0.4! minor changes in almost every section
Andrés, estás ahora como otro corresponding author.
Intro: cambios menores
Resultados: reestructuración de algunas partes. Hemos eliminado los "as discussed in supplementary..." de varias de las partes. Subida la Tabla 1 a su sección. Ampliada la sección de análisis temporal de V y beta.
Discusión: inglés algo más corregido (gracias sobre todo a Jose Manuel). Final menos abrupto y sin tantos detalles innecesarios. Ampliada el párrafo que habla sobre estas "islas de estabilidad" que tan hábilmente encontró JM (a mí me parecen un resulta

% release codenumber (update it here!)
\newcommand\releasecode{0.3.0}

@@

% watermarks (time consuming!)
\newwatermark[allpages,color=PineGreen!4,angle=45,scale=9,xpos=-40,ypos=0]{DRAFT}
\newwatermark[allpages,color=PineGreen!12,angle=0,scale=1,xpos=0,ypos=137]{
\emph{mSystems} \LaTeX\ manuscript release \releasecode}

@@

%TC:newcounter abswords Words in abstract
%TC:macro \abstract [abswords]
{\abstract{Animal microbiota (human included) plays an important role keeping healthy the physiological status of the host. Increasing research activity is dedicated to understand how changes in composition and function of the microbiota are associated to disease or not. We analyze 16S rRNA and whole genome sequencing (WGS) published data from the gut microbiota of 97 individuals monitored in time. Temporal fluctuations in the microbial composition reveal significant differences due to factors such us dietary changes, antibiotic intake, age or disease. Here we show that a fluctuation scaling law describes the temporal changes in the gut microbiota. This law allows to estimate the temporal variability of the microbial population and quantitatively characterizes the path toward disease by a noise-induced phase transition. The estimation of the systemic parameters for follow-up studies may have clinical use and, more generally, applications in other fields where it is important to know if a given community is stable or not.}}

@@

%===================================== SECTIONS =======================================
%%% Introduction
\section*{Introduction}
\input{mSys_intro.tex}
\clearpage
%%% Result
\section*{Results}
\input{mSys_results.tex}
\clearpage
%%% Discussion
\section*{Discussion}
\input{mSys_discussion.tex}
\clearpage    
%%% Material & Methods
\section*{Materials and Methods}
\input{mSys_MandM.tex}   
\clearpage

@@

@@

\bibitem{cohen_bac} {\bf Kaltz O, Escobar-Paramo P, Hochberg M, Cohen} JE. 2012. Bacterial microcosms obey Taylor’s law: Effects of abiotic and biotic stress and genetics on mean and variance of population density. Ecol Process {\bf 1}:5.

@@

\bibitem{QIIME} {\bf Caporaso JG, Kuczynski J, Stombaugh J, Bittinger K, Bushman FD, Costello EK, Fierer N, Peña AG, Goodrich JK, Gordon JI, Huttley G a, Kelley ST, Knights D, Koenig JE, Ley RE, Lozupone C a, Mcdonald D, Muegge BD, Pirrung M, Reeder J, Sevinsky JR, Turnbaugh PJ, Walters W a, Widmann J, Yatsunenko T, Zaneveld J, Knight R.} 2010. correspondence QIIME allows analysis of high- throughput community sequencing data Intensity normalization improves color calling in SOLiD sequencing. Nat Publ Gr {\bf 7}:335–336.
\bibitem{LMAT} {\bf Ames SK, Hysom DA, Gardner SN, Lloyd GS, Gokhale MB, Allen JE}.  2013. Scalable metagenomic taxonomy classification using a reference genome database.  Bioinformatics {\bf 29}:2253-2260.
\bibitem{LMAT2} {\bf Ames SK, Gardner SN, Marti JM, Slezak TR, Gokhale MB, Allen JE}. 2015. Using populations of human and microbial genomes for organism detection in metagenomes. Genome Res. {\bf 25}:1056-67.
\bibitem{fs} {\bf Eisler Z, Bartos I, Kertész J.} 2008. Fluctuation scaling in complex systems: Taylor’s law and beyond1. Adv Phys {\bf 57}:89–142.
\bibitem{stat} {\bf Jørgensen B, Martinez JR, Tsao M.} 1994. Asymptotic behaviour of the variance function. Scand J Stat {\bf 21}:223–243.

@@

%discussion
\bibitem {kilpatrick} {\bf Kilpatrick a M, Ives a R.} 2003. Species interactions can explain Taylor’s power law for ecological time series. Nature {\bf 422}:65–68.
\bibitem {ballantyne} {\bf Ballantyne IV F, J. Kerkhoff A.} 2007. The observed range for temporal mean-variance scaling exponents can be explained by reproductive correlation. Oikos {\bf 116}:174–180.
\bibitem {joao} {\bf Stein RR, Bucci V, Toussaint NC, Buffie CG, Rätsch G, Pamer EG, Sander C, Xavier JB.}  2013. Ecological modeling from time-series inference: insight into dynamics and stability of intestinal microbiota. PLoS Comput Biol {\bf 9}:e1003388.
\bibitem {mehta} {\bf Fisher CK, Mehta P.} 2014. Identifying keystone species in the human gut microbiome from metagenomic timeseries using sparse linear regression. PLoS One {\bf 9}:e102451.
\bibitem {bucci}  {\bf Bucci V, Tzen B, Li N, Simmons M, Tanoue T, Bogart E, Deng L, Yeliseyev V, Delaney ML, Liu Q, Olle B, Stein RR, Honda K, Bry L, Gerber GK.} 2016. MDSINE: Microbial Dynamical Systems INference Engine for microbiome time-series analyses. Genome Biol {\bf 17}:121.
\bibitem {cohen_taylor} {\bf Cohen JE, Xu M, Schuster WSF}. 2013. Stochastic multiplicative population growth predicts and interprets Taylor ’ s power law of fluctuation scaling Stochastic multiplicative population growth predicts and interprets Taylor ’ s power law of fluctuation scaling. Proc R Soc B Biol Sci {\bf 280}:20122955.
\bibitem {koenig} {\bf Koenig JE, Spor A, Scalfone N, Fricker AD, Stombaugh J, Knight R, Angenent LT, Ley RE.} 2011. Succession of microbial consortia in the developing infant gut microbiome. Proc Natl Acad Sci {\bf 108} :4578–4585.
\bibitem {tikhonov} {\bf  Tikhonov M.} 2016. Community-level cohesion without cooperation. Elife {\bf 5}.  \todo{Revisar referencias a partir de aquí: si proceden una vez que CC no se incluye y, en caso aformativo, adaptar el formato}
	\bibitem{FD} Weber, J. \textit{et al.} Fluctuation dissipation theorem. {\it Phys. Rev.} {\bf 101}, 1620-6 (1956).
	\bibitem{FASTX} Gordon, A., Hannon, G.J. FASTX-Toolkit. FASTQ/A shortreads pre-processing tools (2010). http://hannonlab.cshl.edu/fastx\_toolkit/ (accessed 23 Feb 2015).
	\bibitem{SILVA} Quast C.  \textit{et al.} The SILVA ribosomal RNA gene database project: improved data processing and web-based tools (2013)
 	\bibitem{ecology} Xiao Xiao, Ethan P. White, Mevin B. Hooten, and Susan L. Durham. On the use of log-transformation vs. nonlinear regression for analyzing biological power laws. {\it Ecology} {\bf 92}, 10, 1887-1894 (2011).
 	\bibitem{genR2} Magee L., $R^2$ measures based on wald and likelihood ratio joint significance tests. {\it The American Statistician} {\bf 44}, 3, 250-253 (1990).
	\bibitem{disR2} Nagelkerke N.J.D., A note on a general definition of the coefficient of determination. {\it Biometrika} {\bf 78}, 3, 691-692 (1991).
	\bibitem{boot} Wu, C.F.J. Jackknife, bootstrap and other resampling methods in regression analysis. (with discussions) \textit{The Annals of Statistics} {\bf 14}: 1261?1350 (1986)
\end{thebibliography}

>>>>

% release codenumber (update it here!)
\newcommand\releasecode{0.5.0}

@@

% watermarks (time consuming!)
\newwatermark[allpages,color=Fuchsia!4,angle=45,scale=9,xpos=-40,ypos=0]{DRAFT}
\newwatermark[allpages,color=Fuchsia!12,angle=0,scale=1,xpos=0,ypos=137]{
\emph{mSystems} \LaTeX\ manuscript release \releasecode}

@@

%TC:newcounter abswords Words in abstract
%TC:macro \abstract [abswords]
{\abstract{Animal microbiota (human included) plays an important role keeping healthy the physiological status of the host. Increasing research activity is dedicated to understand how changes in composition and function of the microbiota are associated to disease or not. We analyze 16S rRNA and shotgun metagenomic sequencing (SMS) published data from the gut microbiota of 97 individuals monitored in time. Temporal fluctuations in the microbial composition reveal significant differences due to factors such us dietary changes, antibiotic intake, age or disease. Here we show that a fluctuation scaling law describes the temporal changes in the gut microbiota. This law allows to estimate the temporal variability of the microbial population and quantitatively characterizes the path toward disease by a noise-induced phase transition. The estimation of the systemic parameters for follow-up studies may have clinical use and, more generally, applications in other fields where it is important to know if a given community is stable or not.}}

@@

%===================================== SECTIONS =======================================
%%% Introduction
\section*{Introduction}
\input{mSys_intro.tex}
%\clearpage
%%% Result
\section*{Results}
\input{mSys_results.tex}
%\clearpage
%%% Discussion
\section*{Discussion}
\input{mSys_discussion.tex}
%\clearpage    
%%% Material & Methods
\section*{Materials and Methods}
\input{mSys_MandM.tex}   
%\clearpage

@@

\bibitem{CFS} {\bf Giloteaux L, Goodrich JK, Walters WA, Levine SM, Ley RE, Hanson MR}. 2016.
Reduced diversity and altered composition of the gut microbiome in individuals with myalgic encephalomyelitis/chronic fatigue syndrome. Microbiome. {\bf 4}:30

@@

\bibitem{cohen_bac} {\bf Kaltz O, Escobar-Paramo P, Hochberg M, Cohen} JE. 2012. Bacterial microcosmos obey Taylor’s law: Effects of abiotic and biotic stress and genetics on mean and variance of population density. Ecol Process {\bf 1}:5.

@@

\bibitem{FASTX} {\bf Gordon, A., Hannon, G.J.} 2010. FASTX-Toolkit. FASTQ/A shortreads pre-processing tools v0.0.13. http://hannonlab.cshl.edu/fastx\_toolkit/ (last accessed 26 Jul 2016).
\bibitem{QIIME} {\bf Caporaso JG, Kuczynski J, Stombaugh J, Bittinger K, Bushman FD, Costello EK, Fierer N, Peña AG, Goodrich JK, Gordon JI, Huttley G a, Kelley ST, Knights D, Koenig JE, Ley RE, Lozupone C a, Mcdonald D, Muegge BD, Pirrung M, Reeder J, Sevinsky JR, Turnbaugh PJ, Walters W a, Widmann J, Yatsunenko T, Zaneveld J, Knight R.} 2010. correspondence QIIME allows analysis of high- throughput community sequencing data Intensity normalization improves color calling in SOLiD sequencing. Nat Publ Gr {\bf 7}:335–336.
\bibitem{SILVA} {\bf Quast C, Pruesse E, Yilmaz P, Gerken J, Schweer T, Yarza P, Peplies J, Glöckner FO}. 2013. The SILVA ribosomal RNA gene database project: improved data processing and web-based tools. Acids Res. {\bf 41}:D590-D596
\bibitem{LMAT} {\bf Ames SK, Hysom DA, Gardner SN, Lloyd GS, Gokhale MB, Allen JE}.  2013. Scalable metagenomic taxonomy classification using a reference genome database.  Bioinformatics {\bf 29}:2253-2260.
\bibitem{LMAT2} {\bf Ames SK, Gardner SN, Marti JM, Slezak TR, Gokhale MB, Allen JE}. 2015. Using populations of human and microbial genomes for organism detection in metagenomes. Genome Res. {\bf 25}:1056-67.
\bibitem{fs} {\bf Eisler Z, Bartos I, Kertész J.} 2008. Fluctuation scaling in complex systems: Taylor’s law and beyond1. Adv Phys {\bf 57}:89–142.
\bibitem{stat} {\bf Jørgensen B, Martinez JR, Tsao M.} 1994. Asymptotic behaviour of the variance function. Scand J Stat {\bf 21}:223–243.

@@

%discussion
\bibitem {kilpatrick} {\bf Kilpatrick a M, Ives a R.} 2003. Species interactions can explain Taylor’s power law for ecological time series. Nature {\bf 422}:65–68.
\bibitem {ballantyne} {\bf Ballantyne IV F, J. Kerkhoff A.} 2007. The observed range for temporal mean-variance scaling exponents can be explained by reproductive correlation. Oikos {\bf 116}:174–180.
\bibitem {joao} {\bf Stein RR, Bucci V, Toussaint NC, Buffie CG, Rätsch G, Pamer EG, Sander C, Xavier JB.}  2013. Ecological modeling from time-series inference: insight into dynamics and stability of intestinal microbiota. PLoS Comput Biol {\bf 9}:e1003388.
\bibitem {mehta} {\bf Fisher CK, Mehta P.} 2014. Identifying keystone species in the human gut microbiome from metagenomic timeseries using sparse linear regression. PLoS One {\bf 9}:e102451.
\bibitem {bucci}  {\bf Bucci V, Tzen B, Li N, Simmons M, Tanoue T, Bogart E, Deng L, Yeliseyev V, Delaney ML, Liu Q, Olle B, Stein RR, Honda K, Bry L, Gerber GK.} 2016. MDSINE: Microbial Dynamical Systems INference Engine for microbiome time-series analyses. Genome Biol {\bf 17}:121.
\bibitem {cohen_taylor} {\bf Cohen JE, Xu M, Schuster WSF}. 2013. Stochastic multiplicative population growth predicts and interprets Taylor ’ s power law of fluctuation scaling Stochastic multiplicative population growth predicts and interprets Taylor ’ s power law of fluctuation scaling. Proc R Soc B Biol Sci {\bf 280}:20122955.
\bibitem {koenig} {\bf Koenig JE, Spor A, Scalfone N, Fricker AD, Stombaugh J, Knight R, Angenent LT, Ley RE.} 2011. Succession of microbial consortia in the developing infant gut microbiome. Proc Natl Acad Sci {\bf 108}:4578–4585.
\bibitem {FD} {\bf Weber, J.} 1956. Fluctuation Dissipation Theorem. Phys. Rev. {\bf 101}:1620-6
\bibitem {tikhonov} {\bf  Tikhonov M.} 2016. Community-level cohesion without cooperation. Elife {\bf 5}.

%M&M
\bibitem{biopwrlaw} {\bf Xiao X, White EP, Hooten MB, Durham SL}. 2011. On the use of log-transformation vs. nonlinear regression for analyzing biological power laws. Ecology {\bf 92}:1887-1894
\bibitem{boot} {\bf Wu, C.F.J.} 1986. Jackknife, bootstrap and other resampling methods in regression analysis. (with discussions) \textit{The Annals of Statistics} {\bf 14}:1261-1350 
\end{thebibliography}

### 2016-07-26 mSystems/mSys_MandM.tex
Several minor changes leading to 0.5.0, another step towards pre-releasing
On branch master changes to be committed:
	modified:   mSys_MAIN.pdf
	modified:   mSys_MAIN.tex
	modified:   mSys_MandM.tex
	modified:   mSys_discussion.tex
	modified:   mSys_intro.tex
	modified:   mSys_results.tex

We model the microbial abundances across time along the lines of Blumm \textit{et al.}\cite{ranking}. The dynamics of taxon relative abundances is described by the Langevin equation:
\begin{linenomath}
\begin{equation}
\dot{x_i} = F_i \cdot x_i^\alpha + V \cdot x_i^\beta \xi_i(t) - \phi(t) \cdot x_i,
\end{equation}
\end{linenomath}
where F$_i$ captures the fitness of the taxon i, V corresponds to the noise amplitude and $\xi_i$(t) is a Gaussian random noise with zero mean  $<\xi_i(t)>$ = 0 and variance uncorrelated in time, $<\xi_i(t) \xi_i(t')>$ =  $\delta(t'-t)$, . The function $\phi(t)$ ensures the normalization at all times, $\sum x_i(t) = 1$, and corresponds to $\phi(t) = \sum F_i x_i^\alpha + \sum V x_i^\beta \xi_i(t)$.
The temporal evolution of the probability that a taxon i has a relative abundance $x_i(t)$, P(x$_i$,t), is determined by the Fokker-Planck equation:
\begin{linenomath}
\begin{equation}
\frac{\partial P}{\partial t} = - \frac{\partial}{\partial x_i}  [(F_i \cdot x_i^\alpha - \phi(t) \cdot x_i ) \cdot P]+ \frac{1}{2} \frac{\partial^2}{\partial x_i^2} (V^2 \cdot x_i^{2\beta}\cdot P).
\end{equation}
\end{linenomath}
The microbiota evolves towards a steady-state with a time-independent probability depending on the values of $\alpha$, $\beta$, F$_i$ and V. For $\alpha<1$ (otherwise, systems are always unstable), the steady-state probability may be localized in a region around a preferred value or broadly distributed over a wide range, depending on whether the fitness F$_i$  dominates or is overwhelmed by the noise amplitude V. The steady-state solution of the Fokker-Planck equation is given by:
\begin{eqnarray}
P_0 (x_i) &=& C_{ne}(\alpha,\beta,F_i,V)  \cdot x_i^{-2\beta}  \cdot \exp\Big[\frac{2F_i}{V^2}\frac{x_i^{1+\alpha-2\beta}}{1+\alpha-2\beta}-\frac{\phi_0}{V^2}\frac{x_i^{2-2\beta}}{1-\beta}\Big] \quad \textrm{if} \quad  2\beta \ne 1+\alpha, \notag \\
P_0 (x_i) &=& C_e(\alpha,\beta,F_i,V)  \cdot x_i^{\frac{2F_i}{V^2} -2\beta}  \cdot \exp\Big[\frac{\phi_0}{V^2}\frac{x_i^{2-2\beta}}{1-\beta}\Big] \quad \textrm{if} \quad  2\beta = 1+\alpha,
\notag
\end{eqnarray}
where $\phi_0 = (\sum_i F_i^{1/(1-\alpha)})^{1-\alpha}$ and C$_{ne}$ and C$_{e}$ are integrals that should be solved numerically for the parameters of interest. The ordered phase happens when the solution has a maximum in the physical interval ($0<x_i<1$). For larger V, the transition to a disordered phase happens when the maximum shifts to the unphysical region $x_i<0$, which sets the phase transition region V($\alpha,\beta,F_i$). The phase transition region can be calculated analytically in particular cases:
\begin{eqnarray}
F_i^2 &=& 4 \beta \phi_0 V^2\quad \textrm{if} \quad  \beta = \alpha \neq 1, \notag \\
F_i &=& \beta V^2\quad \textrm{if} \quad  2\beta = 1+\alpha,
\notag
\end{eqnarray}
where the first case, simplifies to $F= 3 V^2$ if $\beta = 0.75$ and the fitness of this taxon dominates in $\phi_0$. 
In many physical systems (Brownian motion is the classical example), the two terms of the Langevin equation are related.  The \emph{fluctuation--dissipation theorem} states a general relationship between the response to an external disturbance and the internal fluctuations of the system\cite{FD}. The theorem can be used as the basic formula to derive the fitness from the analysis of fluctuations 
of the microbiota, assuming that it is in equilibrium (the ordered phase).

\todo{Explain better the fluctuation-dissipation theorem}

@@

%%
\subsubsection*{Sample selection}
We have chosen studies about relevant pathologies containing metagenomic sequencing time data series of bacterial populations from humans in different healthy and non-healthy states. We have selected only those individuals who had three or more time points of data available in databases. Caporaso et al. study \cite{moving} was selected as it has two healthy individuals measured over a very large timespan, with almost daily sampling. Faith et al. study \cite{LEA} was selected due to the BMI differences between subjects, moreover some of them had diets which can be treated as system perturbations. We considered healthy only those individuals who had normal or overweight BMI. Smith et al., study \cite{kwashiorkor} was selected for both the age of the patients and the rare disease. We only worked with the discordant twins, and considered healthy those who were not affected by kwashiorkor in each pair of patients. David et al., study \cite{diet} was selected for its differential diets. We considered as the healthy part the first time samples of each individual before the diet, and the rest of time points as perturbations. Dethlefsen and Relman work \cite{antibiotics} was selected due to the interesting treatment of two antibiotic intakes of the same antibiotic by three different subjects. We considered as the healthy part only those times before any antibiotic treatment, and as perturbations the time of antibiotic intakes and the period after that. And finally, we also considered a study made in our group carried by Durban et al., \cite{IBS} in which we considered as healthy subjects those who didn't suffer from irritable bowel disease, and as perturbation the patients who had this disease.

Metadata of each study is provided in Tables \ref{tab:diet} to \ref{tab:LEA}. All used 16S rRNA gene sequencing except for the study of the discordant kwashiorkor twins\cite{kwashiorkor} (see Tables \ref{tab:DH} and \ref{tab:DK}) where shotgun metagenomic sequencing (SMS) and 16S rRNA were used. In the latter case we selected to work with SMS data to show that our method is valid regardless of the source of taxonomic information. Each one of the datasets was treated as follows:

@@

%%
\subsubsection*{Metagenomic sequences processing}
Metagenomic shotgun (and 16S too) sequences were analyzed with LMAT (Livermore Metagenomics Analysis Toolkit) software package\cite{LMAT} (version 1.2.4, with Feb'15 release of data base \emph{LMAT-Grand}). LMAT was run using a Bull shared-memory node belonging to the team's HPC (high performance computing) cluster. It is equipped with 32 cores (64 threads available using Intel Hyper-threading technology) as it has two Haswell-based Xeons ($22\un{nm}$ technology), the E5-2698v3@2.3 GHz, sharing half a tebibyte (0.5 TiB, that is, 512 gibibytes) of DRAM memory. This node is also provided with a PCIe SSD card as NVRAM, the Micron P420m HHHL, with 1.4 TB, and 750000 reading IOPS, 4 KB, achieving 3.3 GB/s. The computing node was supplied with a RAID-0 (striping) scratch disk area. We used the ``Grand'' database\cite{LMAT2}, release Feb'15, provided by the LMAT team, where ``Grand'' refers to a huge database that contains k-mers from all viral, prokaryote, fungal and protist genomes present in the NCBI database, plus Human reference genome (hg19), plus GenBank Human, plus the 1000 Human Genomes Project (HGP) (this represent about 31.75 billion k-mers occupying 457.62 GB)\cite{LMAT2}. Previously to any calculation, the full database was loaded in the NVRAM. With this configuration the observed LMAT sustained sequence classification rate was $20\unclassrate$. Finally, it is worth mentioning that a complete set of Python scripts have been developed as back-end and front-end of the LMAT pipeline in order to manage the added complexity of time series analysis.

@@

The basic idea of bootstrapping is that inference about a population from sample data (sample $\rightarrow$ population) can be modeled by resampling the sample data and performing inference on (resample $\rightarrow$ sample). To adapt this general idea to our problem, we resample the x-data array using its errors array. That is, for each replicate, a new x-data array is computed based on:
\begin{linenomath}
$$x^*_i = x_i + v_i$$
\end{linenomath}
where $v_i$ is a Gaussian random variable with mean $\mu_i=0$ and standard deviation $\sigma_i=\mathrm{SEM}_i$, as defined previously. For each replicate a complete un-weighted power-law fit is performed, where to choose between fitting power laws ($y=Vx^\beta$) using linear regression on log-transformed (LLR) data versus non-linear regression (NLR) we mainly follow \emph{General Guidelines for the Analysis of Biological Power Laws}\cite{ecology}. The parameters of the X-weighted fit are then estimated by averaging through all the replicate fits performed, and their errors are estimated by computing the standard deviation also for all the fits. At the end of each step, the relative error is calculated by comparing the fit parameters estimation in the last step with the previous one. Finally, both the coefficient of determination of the fit and the coefficient of correlation between the fit parameters are estimated by averaging.

@@

The Rank Stability Index (RSI) is shown as a percentage in a separate bar on the right of the rank matrix plot shown in Figure \ref{fig:corrank}. The RSI is strictly $1$ for an element whose range never changes over time, and is strictly $0$ for an element whose rank oscillates between the extremes from time to time. So, RSI is calculated, per element, as $1$ less the quotient of the number of true rank hops taken between the number of maximum possible rank hops, all powered to $p$:
\begin{linenomath}
$${\rm RSI} = \left(1-\frac{\text{true rank hops}}{\text{possible rank hops}}\right)^p = \left(1-\frac{D}{(N-1)(t-1)}\right)^p$$
\end{linenomath}
where $D$ is the total of rank hops taken by the studied element, $N$ is the number of elements that have been ranked, and $t$ is the number of time samples. The power index $p=4$ is arbitrarily chosen to increase the resolution in the stable region.

>>>>

We model the microbial abundances across time along the lines of Blumm \textit{et al.}\cite{ranking}. The dynamics of taxon relative abundances is described by the Langevin equation:
\begin{linenomath}
\begin{equation*}
\dot{x_i} = F_i \cdot x_i^\alpha + V \cdot x_i^\beta \xi_i(t) - \phi(t) \cdot x_i,
\end{equation*}
\end{linenomath}
where F$_i$ captures the fitness of the taxon i, V corresponds to the noise amplitude and $\xi_i$(t) is a Gaussian random noise with zero mean  $<\xi_i(t)>$ = 0 and variance uncorrelated in time, $<\xi_i(t) \xi_i(t')>$ =  $\delta(t'-t)$, . The function $\phi(t)$ ensures the normalization at all times, $\sum x_i(t) = 1$, and corresponds to $\phi(t) = \sum F_i x_i^\alpha + \sum V x_i^\beta \xi_i(t)$.
The temporal evolution of the probability that a taxon i has a relative abundance $x_i(t)$, P(x$_i$,t), is determined by the Fokker-Planck equation:
\begin{linenomath}
\begin{equation*}
\frac{\partial P}{\partial t} = - \frac{\partial}{\partial x_i}  [(F_i \cdot x_i^\alpha - \phi(t) \cdot x_i ) \cdot P]+ \frac{1}{2} \frac{\partial^2}{\partial x_i^2} (V^2 \cdot x_i^{2\beta}\cdot P).
\end{equation*}
\end{linenomath}
The microbiota evolves towards a steady-state with a time-independent probability depending on the values of $\alpha$, $\beta$, F$_i$ and V. For $\alpha<1$ (otherwise, systems are always unstable), the steady-state probability may be localized in a region around a preferred value or broadly distributed over a wide range, depending on whether the fitness F$_i$  dominates or is overwhelmed by the noise amplitude V. The steady-state solution of the Fokker-Planck equation is given by:
\begin{eqnarray}
P_0 (x_i) &=& C_{ne}(\alpha,\beta,F_i,V)  \cdot x_i^{-2\beta}  \cdot \exp\Big[\frac{2F_i}{V^2}\frac{x_i^{1+\alpha-2\beta}}{1+\alpha-2\beta}-\frac{\phi_0}{V^2}\frac{x_i^{2-2\beta}}{1-\beta}\Big] \quad \textrm{if} \quad  2\beta \ne 1+\alpha, \notag \\
P_0 (x_i) &=& C_e(\alpha,\beta,F_i,V)  \cdot x_i^{\frac{2F_i}{V^2} -2\beta}  \cdot \exp\Big[\frac{\phi_0}{V^2}\frac{x_i^{2-2\beta}}{1-\beta}\Big] \quad \textrm{if} \quad  2\beta = 1+\alpha,
\notag
\end{eqnarray}
where $\phi_0 = (\sum_i F_i^{1/(1-\alpha)})^{1-\alpha}$ and C$_{ne}$ and C$_{e}$ are integrals that should be solved numerically for the parameters of interest. The ordered phase happens when the solution has a maximum in the physical interval ($0<x_i<1$). For larger V, the transition to a disordered phase happens when the maximum shifts to the unphysical region $x_i<0$, which sets the phase transition region V($\alpha,\beta,F_i$). The phase transition region can be calculated analytically in particular cases:
\begin{eqnarray}
F_i^2 &=& 4 \beta \phi_0 V^2\quad \textrm{if} \quad  \beta = \alpha \neq 1, \notag \\
F_i &=& \beta V^2\quad \textrm{if} \quad  2\beta = 1+\alpha,
\notag
\end{eqnarray}
where the first case, simplifies to $F= 3 V^2$ if $\beta = 0.75$ and the fitness of this taxon dominates in $\phi_0$. 
In many physical systems (Brownian motion is the classical example), the two terms of the Langevin equation are related.  The \emph{Fluctuation Dissipation Theorem} states a general relationship between the response to an external disturbance and the internal fluctuations of the system\cite{FD}. The theorem can be used as the basic formula to derive the fitness from the analysis of fluctuations 
of the microbiota, assuming that it is in equilibrium (the ordered phase).

\todo{Explain better the Fluctuation Dissipation Theorem}

@@

%%
\subsubsection*{Sample selection}
We have chosen studies about relevant pathologies containing metagenomic sequencing time data series of bacterial populations from humans in different healthy and non-healthy states. We have selected only those individuals who had three or more time points of data available in databases. Caporaso et al. study \cite{moving} was selected as it has two healthy individuals measured over a very large timespan, with almost daily sampling. Faith et al. study \cite{LEA} was selected due to the BMI differences between subjects, moreover some of them had diets which can be treated as system perturbations. We considered healthy only those individuals who had normal or overweight BMI. Smith et al., study \cite{kwashiorkor} was selected for both the age of the patients and the rare disease. We only worked with the discordant twins, and considered healthy those who were not affected by kwashiorkor in each pair of patients. David et al., study \cite{diet} was selected for its differential diets. We considered as the healthy part the first time samples of each individual before the diet, and the rest of time points as perturbations. Dethlefsen and Relman work \cite{antibiotic} was selected due to the interesting treatment of two antibiotic intakes of the same antibiotic by three different subjects. We considered as the healthy part only those times before any antibiotic treatment, and as perturbations the time of antibiotic intakes and the period after that. And finally, we also considered a study made in our group carried by Durban et al., \cite{IBS} in which we considered as healthy subjects those who didn't suffer from irritable bowel disease, and as perturbation the patients who had this disease.

Metadata of each study is provided in Supplementary Tables S1 to S4. All used 16S rRNA gene sequencing except for the study of the discordant kwashiorkor twins\cite{kwashiorkor}, where both shotgun metagenomic sequencing (SMS) and 16S rRNA were used. In the latter case we selected to work with SMS data to show that our method is valid regardless of the source of taxonomic information. Each one of the datasets was treated as follows:

@@

%%
\subsubsection*{Metagenomic sequences processing}
Shotgun metagenomic (and 16S too) sequences were analyzed with LMAT (Livermore Metagenomics Analysis Toolkit) software package\cite{LMAT} (version 1.2.4, with Feb'15 release of data base \emph{LMAT-Grand}). LMAT was run using a Bull shared-memory node belonging to the team's HPC (high performance computing) cluster. It is equipped with 32 cores (64 threads available using Intel Hyper-threading technology) as it has two Haswell-based Xeons ($22\un{nm}$ technology), the E5-2698v3@2.3 GHz, sharing half a tebibyte (0.5 TiB, that is, 512 gibibytes) of DRAM memory. This node is also provided with a PCIe SSD card as NVRAM, the Micron P420m HHHL, with 1.4 TB, and 750000 reading IOPS, 4 KB, achieving 3.3 GB/s. The computing node was supplied with a RAID-0 (striping) scratch disk area. We used the ``Grand'' database\cite{LMAT2}, release Feb'15, provided by the LMAT team, where ``Grand'' refers to a huge database that contains k-mers from all viral, prokaryote, fungal and protist genomes present in the NCBI database, plus Human reference genome (hg19), plus GenBank Human, plus the 1000 Human Genomes Project (HGP) (this represent about 31.75 billion k-mers occupying 457.62 GB)\cite{LMAT2}. Previously to any calculation, the full database was loaded in the NVRAM. With this configuration the observed LMAT sustained sequence classification rate was $20\unclassrate$. Finally, it is worth mentioning that a complete set of Python scripts have been developed as back-end and front-end of the LMAT pipeline in order to manage the added complexity of time series analysis.

@@

The basic idea of bootstrapping is that inference about a population from sample data (sample $\rightarrow$ population) can be modeled by resampling the sample data and performing inference on (resample $\rightarrow$ sample)\cite{boot}. To adapt this general idea to our problem, we resample the x-data array using its errors array. That is, for each replicate, a new x-data array is computed based on:
\begin{linenomath}
$$x^*_i = x_i + v_i$$
\end{linenomath}
where $v_i$ is a Gaussian random variable with mean $\mu_i=0$ and standard deviation $\sigma_i=\mathrm{SEM}_i$, as defined previously. For each replicate a complete un-weighted power-law fit is performed, where to choose between fitting power laws ($y=Vx^\beta$) using linear regression on log-transformed (LLR) data versus non-linear regression (NLR) we mainly follow \emph{General Guidelines for the Analysis of Biological Power Laws}\cite{biopwrlaw}. The parameters of the X-weighted fit are then estimated by averaging through all the replicate fits performed, and their errors are estimated by computing the standard deviation also for all the fits. At the end of each step, the relative error is calculated by comparing the fit parameters estimation in the last step with the previous one. Finally, both the coefficient of determination of the fit and the coefficient of correlation between the fit parameters are estimated by averaging.

@@

The Rank Stability Index (RSI) is shown as a percentage in a separate bar on the right of the rank matrix plot shown in Figures \ref{fig:corrank_IBS_A} and \ref{fig:corrank_IBS_P2}. The RSI is strictly $1$ for an element whose range never changes over time, and is strictly $0$ for an element whose rank oscillates between the extremes from time to time. So, RSI is calculated, per element, as $1$ less the quotient of the number of true rank hops taken between the number of maximum possible rank hops, all powered to $p$:
\begin{linenomath}
$${\rm RSI} = \left(1-\frac{\text{true rank hops}}{\text{possible rank hops}}\right)^p = \left(1-\frac{D}{(N-1)(t-1)}\right)^p$$
\end{linenomath}
where $D$ is the total of rank hops taken by the studied element, $N$ is the number of elements that have been ranked, and $t$ is the number of time samples. The power index $p=4$ is arbitrarily chosen to increase the resolution in the stable region.
